\documentclass[conference]{IEEEtran}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{newtxtext,newtxmath}
\usepackage{amsmath,amssymb,bm}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{enumitem}
\usepackage{cite}
\usepackage[breaklinks=true,hidelinks]{hyperref}
\usepackage[nameinlink,capitalise,noabbrev]{cleveref}

% Theorem styles
\newtheoremstyle{bodysmall}
  {3pt}{3pt}{\normalfont\small}{0pt}{\bfseries\small}{.}{0.5em}{}
\theoremstyle{bodysmall}
\newtheorem*{theorem*}{Theorem}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}

\title{A Practical Description of the KalmANF Frequency Tracker\\
and Its C\texttt{++} Implementation}

\author{%
  \IEEEauthorblockN{Mikhail S.\ Grushinskiy}
  \IEEEauthorblockA{Independent Researcher\\
  Bareboat Necessities Project\\
  \texttt{mgrouch@users.github.com}}
}

\begin{document}
\maketitle

% ---------------------------------------------------------------
\begin{abstract}
This note gives a clean and practical description of the KalmANF frequency
tracker as implemented in C\texttt{++}, based on the algorithm of Ali and van
Waterschoot~\cite{AliWaterschoot2023}. The method adapts the coefficient of a
second-order resonator using a scalar Kalman filter, recovering the dominant
sinusoidal frequency sample by sample.

This document describes what the code actually computes at each step, why
each step is structured the way it is, and how the parameters interact. It
corrects several discrepancies between standard textbook presentations and the
actual implementation: in particular, the Jacobian used in the Kalman gain
does not carry a $\rho$ factor, and the innovation is formed without $\rho$
scaling. A brief convergence argument, a discussion of DC offset sensitivity,
and the synthesised confidence and lock layer are also included.
\end{abstract}

\begin{IEEEkeywords}
frequency estimation, adaptive notch filter, Kalman filter, sinusoidal
tracking, embedded signal processing
\end{IEEEkeywords}

% ---------------------------------------------------------------
\section{Introduction}

The KalmANF tracker estimates the instantaneous frequency of a scalar
oscillatory signal by fitting a single adaptive parameter: the coefficient
$a = 2\cos(\hat\omega_d)$ of a second-order digital resonator, where
$\hat\omega_d$ is the estimated digital radian frequency per sample. Once $a$
is known, the physical frequency follows directly.

The adaptation rule is a one-dimensional Kalman filter. The Kalman framework
is attractive here because it provides a principled trade-off between tracking
speed and noise sensitivity, with two intuitive tuning knobs: a process noise
$q$ (how fast can $a$ change?) and a measurement noise $r$ (how much do we
trust the residual?).

Compared to purely gradient-based approaches, the Kalman structure gives
automatic gain scheduling: the effective adaptation step size $K[n]$ shrinks
as the covariance $p$ decreases, so the tracker commits more strongly to its
estimate over time. Reinjecting process noise $q$ at each step prevents the
gain from collapsing to zero and keeps the tracker responsive to genuine
frequency shifts.

The following sections describe each component in detail, explaining both
what is computed and why.

% ---------------------------------------------------------------
\section{Signal Model}

Let $y[n]$ denote the measured signal at sample $n$, with physical sample
period $\Delta t$. In the intended operating regime, $y[n]$ is dominated by
a slowly varying sinusoidal component:
\begin{equation}\label{eq:signal}
  y(t) = A(t)\sin\!\bigl(\omega(t)\,t + \phi(t)\bigr) + \nu(t),
\end{equation}
where $A(t)$ is a slowly varying amplitude, $\omega(t)$ is the instantaneous
angular frequency, $\phi(t)$ is the phase, and $\nu(t)$ is zero-mean noise.
``Slowly varying'' means $|\dot{A}/A| \ll \omega$ and $|\dot\omega| \ll
\omega^2$, so the signal looks locally sinusoidal.

The tracker assumes a single dominant sinusoidal component. Multi-tone signals
will cause the adaptive coefficient $a$ to settle near whichever tone produces
the largest resonator response.

% ---------------------------------------------------------------
\section{Resonator Structure}\label{sec:resonator}

\subsection{Transfer Function}

The core of the tracker is a second-order resonator (also described as a
notch filter from the complementary perspective). Its transfer function is
\begin{equation}\label{eq:tf}
  H(z^{-1};\,a) =
  \frac{1 - a\,z^{-1} + z^{-2}}
       {1 - \rho\,a\,z^{-1} + \rho^2 z^{-2}},
\end{equation}
where $a = 2\cos(\hat\omega_d)$ encodes the tracked digital frequency, and
$\rho \in (0,1)$ is the pole radius.

The numerator places zeros on the unit circle at $e^{\pm j\hat\omega_d}$,
creating a notch at the tracked frequency. The denominator places poles just
inside the unit circle at $\rho\,e^{\pm j\hat\omega_d}$, creating a resonance
(high gain) at the same frequency. The combined effect is a resonator that
rings strongly at $\hat\omega_d$ and attenuates everything else.

\subsection{Resonator State Equation}

Implementing $H$ as a direct-form IIR filter, the internal resonator state
at each sample is
\begin{equation}\label{eq:resonator}
  s[n] = y[n] + \rho\,a\,s[n-1] - \rho^2\,s[n-2].
\end{equation}
This is the denominator recursion of \eqref{eq:tf}: the state $s[n]$
accumulates energy at the resonant frequency. When $a$ matches the true
frequency, the resonator rings in synchrony with the input and its internal
state $s[n-1]$ correlates maximally with future inputs.

\subsection{Frequency Recovery}

Given the current coefficient $a \in (-2, 2)$, the estimated digital
frequency is
\begin{equation}\label{eq:omega_d}
  \hat\omega_d = \arccos\!\left(\frac{a}{2}\right) \quad [\text{rad/sample}],
\end{equation}
and the physical frequency in Hz is
\begin{equation}\label{eq:f_hz}
  \hat{f} = \frac{\hat\omega_d}{2\pi\,\Delta t}.
\end{equation}
The domain constraint $a \in (-2, 2)$ is the condition for $\arccos(a/2)$ to
be real; the code enforces this by clamping $a$ to $(-1.99999, 1.99999)$
after each update.

\emph{Interpretation of $a$.}  At $a \approx 2$ the tracked frequency is
near DC; at $a \approx 0$ it is near $f_s/4$; at $a \approx -2$ it is near
the Nyquist frequency $f_s/2$.

\subsection{Pole Radius and Bandwidth}

The pole radius $\rho$ controls the sharpness of the resonance. A
useful approximation for the $-3\,\text{dB}$ bandwidth is
\begin{equation}\label{eq:bw}
  \mathrm{BW} \approx \frac{1-\rho}{\pi\,\Delta t}.
\end{equation}
Larger $\rho$ gives a narrower resonance, higher frequency resolution, and
longer settling time. Smaller $\rho$ gives a wider resonance, more
robustness to noise and frequency drift, but lower selectivity. A typical
starting point is $\rho = 0.995$.

% ---------------------------------------------------------------
\section{Scalar Kalman Adaptation}\label{sec:kalman}

\subsection{State and Measurement Model}

The tracked parameter $a$ plays the role of the Kalman state. Its evolution
is modelled as a random walk:
\begin{equation}\label{eq:process}
  a[n] = a[n-1] + w[n], \qquad w[n] \sim (0, q),
\end{equation}
where $q > 0$ is the process noise variance. This models the assumption that
$a$ drifts slowly but does not jump discontinuously.

The innovation---the scalar ``measurement'' that drives the Kalman update---is
\begin{equation}\label{eq:innovation}
  e[n] = s[n] - a[n-1]\,s[n-1] + s[n-2].
\end{equation}
\emph{Why this form?}  Comparing with the resonator recursion
\eqref{eq:resonator}, one can write
\begin{equation}
  e[n] = s[n] - \rho\,a\,s[n-1] + \rho^2 s[n-2]
         + (\rho-1)a\,s[n-1] + (1-\rho^2)s[n-2].
\end{equation}
For $\rho \approx 1$ the correction terms are small, and $e[n] \approx y[n]$
is the notch filter output---the part of $y[n]$ that the resonator did
\emph{not} absorb. When $a$ is correct, the resonator absorbs the sinusoidal
component and $e[n]$ contains only noise. This is the signal the Kalman filter
tries to drive toward zero.

The Jacobian (linearised sensitivity of $e$ with respect to $a$, evaluated
at the current $a[n-1]$) is
\begin{equation}\label{eq:jacobian}
  H[n] = \frac{\partial e[n]}{\partial a} = -s[n-1].
\end{equation}
The sign is absorbed into the gain, so the effective observation vector used
in the gain computation is simply $|H[n]| = s[n-1]$.

\begin{remark}
Some presentations of this algorithm write the Jacobian as
$H[n] = -\rho\,s[n-1]$, which would be appropriate if the resonator state
$s[n]$ were differentiated with respect to $a$ while holding $s[n-1]$ fixed.
The implementation uses $H[n] = -s[n-1]$ (without $\rho$), consistent with
the innovation \eqref{eq:innovation} which also has no $\rho$ factor. The two
choices lead to different gain schedules but the same qualitative behaviour.
\end{remark}

\subsection{Kalman Update Equations}

The per-sample Kalman cycle consists of four steps.

\paragraph{1. Prediction.}  Propagate the covariance forward under the
random-walk model:
\begin{equation}\label{eq:pred}
  p_{n|n-1} = p_{n-1} + q.
\end{equation}
Adding $q$ each step prevents the gain from collapsing to zero and keeps the
tracker adaptive. Larger $q$ means faster forgetting of the previous estimate.

\paragraph{2. Kalman gain.}  The optimal gain balances uncertainty in $a$
against noise in $e[n]$:
\begin{equation}\label{eq:gain}
  K[n] = \frac{s[n-1]}{s[n-1]^2 + r\,/\,p_{n|n-1}}.
\end{equation}
This is the standard scalar EKF gain $K = p H^T / (H p H^T + R)$ with
$H = s[n-1]$, $p = p_{n|n-1}$, $R = r$, simplified algebraically.
When the signal is strong ($s[n-1]^2 \gg r/p$), $K \approx 1/s[n-1]$ and the
update normalises by signal amplitude. When the signal is weak or $r$ is
large, $K \to 0$ and the parameter is barely moved.

\paragraph{3. Parameter and covariance update.}
\begin{align}
  a[n] &= a[n-1] + K[n]\,e[n], \label{eq:a_update}\\
  p_n  &= \bigl(1 - K[n]\,s[n-1]\bigr)\,p_{n|n-1}. \label{eq:p_update}
\end{align}
After the update, $a[n]$ is clamped to $(-1.99999,\,1.99999)$ to keep
$\arccos(a/2)$ well-defined, and $p_n$ is floored at $10^{-12}$ to prevent
numerical underflow.

\paragraph{4. Frequency output.}  Convert the updated coefficient to Hz via
\eqref{eq:omega_d}--\eqref{eq:f_hz}.

% ---------------------------------------------------------------
\section{Per-Sample Update Sequence}\label{sec:update}

At each valid sample $(y[n], \Delta t)$, the code executes the following
steps in order. Non-finite inputs or non-positive timesteps are silently
skipped, leaving all states unchanged---a safeguard against NaN propagation.

\begin{enumerate}[leftmargin=1.6em]
  \item Compute the new resonator state from \eqref{eq:resonator},
        using the \emph{current} $a[n-1]$ and previous states $s[n-1]$,
        $s[n-2]$.
  \item Advance the covariance: $p \leftarrow p + q$.
  \item Compute the Kalman gain from \eqref{eq:gain} using the pre-update $p$.
  \item Compute the innovation $e[n]$ from \eqref{eq:innovation}.
  \item Update $a$ via \eqref{eq:a_update} and clamp to $(-1.99999,
        1.99999)$.
  \item Update $p$ via \eqref{eq:p_update} and floor at $10^{-12}$.
  \item Compute $\hat{f}$ from \eqref{eq:omega_d}--\eqref{eq:f_hz}.
  \item Advance resonator history: $s[n-2] \leftarrow s[n-1]$, $s[n-1]
        \leftarrow s[n]$.
  \item Update confidence and lock (\cref{sec:lock}).
\end{enumerate}

The ordering in steps 3--6 matters: the gain must be computed from
$p_{n|n-1}$ (after adding $q$ but before the measurement update), then $p$
is reduced by the measurement. Swapping steps 3 and 4 is safe since $e[n]$
does not depend on $p$.

\begin{table}[t]
  \centering
  \caption{Main variables and their roles}
  \label{tab:vars}
  \small
  \begin{tabular}{@{}llp{3.6cm}@{}}
    \toprule
    Symbol & Range & Meaning \\
    \midrule
    $s[n]$       & $\R$       & Resonator state (IIR output) \\
    $a[n]$       & $(-2,2)$   & Adaptive notch coefficient \\
    $p[n]$       & $>0$       & Kalman covariance on $a$ \\
    $e[n]$       & $\R$       & Innovation (notch residual) \\
    $K[n]$       & $\R$       & Kalman gain \\
    $\hat\omega_d$ & $(0,\pi)$ & Digital frequency (rad/sample) \\
    $\hat{f}$    & $>0$       & Physical frequency (Hz) \\
    \bottomrule
  \end{tabular}
\end{table}

% ---------------------------------------------------------------
\section{Initialization}\label{sec:init}

\subsection{Default Initialization}

The default constructor sets $\rho = 0.995$, $q = 10^{-5}$, $r = 10^3$,
$p_0 = 1$, and $a_0 = 1.9999$ (corresponding to a near-DC initial guess).
The resonator history is zeroed: $s[-1] = s[-2] = 0$.

\subsection{Frequency-Seeded Initialization}

A more practical starting point is to seed $a_0$ from a frequency guess:
\begin{equation}
  a_0 = 2\cos(2\pi\,f_{\rm guess}\,\Delta t).
\end{equation}
The \texttt{initFromFreqGuess} method computes this directly. A good guess
reduces convergence time significantly, because the Kalman gain starts large
($p_0$ is large relative to $r$) and shrinks quickly once $a$ is near its
correct value. Starting far from the true frequency means the tracker must
``sweep'' across parameter space, which takes many samples at the default
gain schedule.

\subsection{Tuning $q$ and $r$}

The ratio $q/r$ (not their absolute values) primarily determines the
steady-state gain and tracking bandwidth. Rough guidelines:
\begin{itemize}[leftmargin=1.4em]
  \item \emph{Slow, stationary signal}: $q = 10^{-6}$, $r = 10^3$. Very
        smooth frequency output, slow to respond to genuine shifts.
  \item \emph{Moderate tracking}: $q = 10^{-5}$, $r = 10^3$ (default).
        Good balance for ocean-wave applications.
  \item \emph{Fast, noisy environment}: $q = 10^{-4}$, $r = 10^2$. Faster
        adaptation but noisier frequency output.
\end{itemize}
Increasing $r$ is equivalent to saying ``I don't trust the innovation
$e[n]$''; increasing $q$ is equivalent to saying ``I expect $a$ to change
quickly.''

% ---------------------------------------------------------------
\section{Stability and Convergence}\label{sec:stability}

\subsection{Steady-State Covariance and Gain}

In steady state, if the signal is stationary and $s[n-1]^2 \approx S^2$ on
average, the covariance equation \eqref{eq:pred}--\eqref{eq:p_update} has
a fixed-point satisfying $p^\star = p^\star + q - K^\star p^\star$, which
gives
\begin{equation}\label{eq:ss_p}
  p^\star = \frac{r}{S^2}\!\left(
    \sqrt{1 + \frac{q S^2}{r}} - 1
  \right)^{-1}
  \approx \sqrt{\frac{qr}{S^2}}
  \quad(q S^2 \ll r).
\end{equation}
The steady-state Kalman gain is then $K^\star \approx \sqrt{q/r}/S$, and the
expected squared error on $a$ is
\begin{equation}\label{eq:ss_err}
  \E[(a[n] - a_{\rm true})^2] \;\to\; p^\star \approx \sqrt{qr}/S.
\end{equation}
This shows that reducing $q$ (slower forgetting) and increasing $r$ (more
conservative updates) both reduce the steady-state variance on $a$, at the
cost of slower tracking.

\subsection{Convergence under Persistent Excitation}

\begin{theorem*}
If the input is persistently exciting ($S^2 = \E[s[n-1]^2] > 0$ on average)
and the true frequency is constant, the Kalman update
\eqref{eq:a_update}--\eqref{eq:p_update} drives $a[n] \to a_{\rm true}$
in mean square as $n\to\infty$.
\end{theorem*}

\emph{Argument.}  The Kalman recursion is a stochastic gradient descent with
automatic step-size scheduling. At each step,
\begin{equation}
  a[n] - a_{\rm true}
  = (1 - K[n] s[n-1])(a[n-1] - a_{\rm true}) + K[n]\,v[n],
\end{equation}
where $v[n]$ collects the noise-driven part of $e[n]$.  Because
$0 < 1 - K s < 1$ (the covariance update contracts $p$ toward zero), the
first term contracts geometrically and the second is zero-mean, so the error
variance converges to the steady-state value \eqref{eq:ss_err}.

\begin{remark}
The persistent excitation condition fails if $y[n] \equiv 0$, in which case
$s[n-1] \to 0$ and $K[n] \to 0$, freezing adaptation. This is the correct
behaviour: with no signal there is no information about frequency.
\end{remark}

% ---------------------------------------------------------------
\section{Sensitivity to DC Offsets}\label{sec:dc}

The Kalman update minimises the innovation $e[n]$, which from
\eqref{eq:innovation} is formed from the resonator state $s[n]$ and
its history. If the input $y[n]$ carries a DC offset $y_0$, the resonator
state accumulates a DC component proportional to $y_0 / (1 - \rho a + \rho^2)$.

This biases $e[n]$ systematically, driving $a$ away from the true value.
The fractional frequency error is of order $y_0/A$ where $A$ is signal
amplitude. The algorithm does \textbf{not} inherently reject DC offsets.

\textbf{Recommended mitigation} (in order of simplicity):
\begin{itemize}[leftmargin=1.4em]
  \item \emph{High-pass filter} the input before the tracker. A single
        first-order IIR with cutoff well below the minimum tracked frequency
        suffices.
  \item \emph{Mean subtraction}: maintain a slow running average of $y[n]$
        and subtract it.
  \item \emph{Augment the state}: add an estimated DC level as a second
        Kalman state and subtract it from $y[n]$ at each step.
\end{itemize}
For accelerometer-based heave sensors, where gravity bias or tilt can
introduce a large quasi-DC component, the high-pass pre-filter is strongly
recommended.

% ---------------------------------------------------------------
\section{EKF Interpretation}\label{sec:ekf}

The algorithm is a scalar Extended Kalman Filter (EKF) with:
\begin{itemize}[leftmargin=1.4em]
  \item \emph{State}: $a[n]$, the notch coefficient.
  \item \emph{Process model}: random walk with variance $q$.
  \item \emph{Measurement function}: $h(a) = 0$ (we want the notch output
        $e[n]$ to be zero at the true $a$).
  \item \emph{Linearised observation}: $H[n] = s[n-1]$ (the Jacobian of
        the innovation with respect to $a$, dropping sign).
  \item \emph{Measurement noise}: variance $r$.
\end{itemize}

The ``extended'' part is that the Jacobian $H[n] = s[n-1]$ changes each
sample because $s[n-1]$ depends on the past trajectory of $y$ and $a$. This
is why the filter must recompute $K[n]$ at every step rather than precomputing
a fixed gain.

The scalar structure means the update costs exactly three multiplications and
two additions per sample beyond the resonator recursion. This makes it
suitable for microcontroller deployments at high sample rates.

% ---------------------------------------------------------------
\section{Confidence and Lock Outputs}\label{sec:lock}

The Kalman adaptation produces a frequency estimate but no explicit lock
indicator. The implementation adds a three-component confidence layer.

\subsection{Signal RMS}

Instantaneous signal power, blended from the raw input and the resonator
state, is smoothed by a one-pole filter:
\begin{align}
  P_{\rm sig}[n]
  &\leftarrow P_{\rm sig} + \alpha(\Delta t, \tau_{\rm rms})
    \bigl(y^2 + 0.5\,s[n-1]^2 - P_{\rm sig}\bigr),\\
  \sigma_{\rm rms} &= \sqrt{\max(P_{\rm sig}, 0)},
\end{align}
where $\alpha(\Delta t, \tau) = \Delta t / (\tau + \Delta t)$ is the one-pole
coefficient. Including $s[n-1]$ in the power estimate means confidence
reflects how actively the resonator is engaged, not just raw input amplitude.

\subsection{Error (Residual) RMS}

The innovation $e[n]$ is also smoothed:
\begin{equation}
  P_e \leftarrow P_e + \alpha(\Delta t, \tau_e)(e^2 - P_e),
  \qquad
  \sigma_e = \sqrt{\max(P_e, 0)}.
\end{equation}
A small residual relative to the signal level indicates that the resonator is
absorbing the sinusoidal component well---a sign of good lock.

\subsection{Frequency Stability}

A slow one-pole smoother tracks the frequency trend:
\begin{equation}
  \bar{f} \leftarrow \bar{f} + \alpha(\Delta t, \tau_f)(\hat{f} - \bar{f}).
\end{equation}
The frequency stability score is
\begin{equation}
  r_{\rm stab} = 1 - \min\!\left(1,\;
    \frac{|\hat{f} - \bar{f}|}{0.25\,\max(\bar{f},\,f_{\min})}
  \right),
\end{equation}
which equals 1 when the frequency is perfectly stable and falls toward 0 as
it wanders. The frequency-proportional denominator applies a coarser tolerance
at high frequencies.

\subsection{Confidence Score}

If $\hat{f} \in [f_{\min}, f_{\max}]$, three sub-scores are combined:
\begin{align}
  r_{\rm amp}
    &= \operatorname{clamp}_{[0,1]}\!\left(
       \frac{\sigma_{\rm rms}}{\max(\sigma_{\rm rms,\min},\,10^{-9})}
       \right),\\
  r_{\rm err}
    &= 1 - \min\!\left(1,\;
       \frac{\sigma_e / \max(\sigma_{\rm rms},\,10^{-9})}
            {\gamma}\right),
\end{align}
where $\gamma$ is a soft threshold for the relative residual (default $1.25$).
The target confidence is then
\begin{equation}\label{eq:conf}
  c^\star = r_{\rm amp}\,\bigl(0.20 + 0.45\,r_{\rm err} + 0.35\,r_{\rm stab}\bigr),
\end{equation}
and zero when $\hat{f}$ is outside the valid range.

\emph{Interpretation of the weights.}  The base term $0.20$ ensures that a
strong, quiet-residual signal always contributes positively even with some
frequency wander. The larger weight on $r_{\rm err}$ ($0.45$) reflects that
a small notch residual is the most direct evidence of good frequency tracking.
The stability term ($0.35$) rewards temporal consistency.

Confidence is itself smoothed and clamped:
\begin{equation}
  c \leftarrow \operatorname{clamp}_{[0,1]}\!\bigl(
  c + \alpha(\Delta t, \tau_c)(c^\star - c)\bigr).
\end{equation}

\subsection{Lock Flag}

Lock requires all three gates to pass simultaneously:
\begin{align}
  &\hat{f} \in [f_{\min},\,f_{\max}],\\
  &\sigma_{\rm rms} \ge \sigma_{\rm rms,\min},\\
  &c \ge c_{\rm lock}.
\end{align}
These check, respectively, that the frequency is plausible, the signal is
detectable, and the estimate has been stable long enough. A strong signal
at the wrong frequency, or a correct-frequency but weak signal, will not set
lock.

\begin{table}[t]
  \centering
  \caption{Confidence and lock parameters with defaults}
  \label{tab:lock}
  \small
  \begin{tabular}{@{}lll@{}}
    \toprule
    Parameter & Default & Role \\
    \midrule
    $\tau_{\rm rms}$          & $\SI{8}{s}$   & Signal RMS smoother \\
    $\tau_e$                  & $\SI{5}{s}$   & Error RMS smoother \\
    $\tau_f$                  & $\SI{3}{s}$   & Frequency smoother \\
    $\tau_c$                  & $\SI{6}{s}$   & Confidence smoother \\
    $\sigma_{\rm rms,\min}$   & $0.012$       & Minimum RMS for lock \\
    $c_{\rm lock}$            & $0.65$        & Confidence threshold \\
    $\gamma$                  & $1.25$        & Relative residual threshold \\
    $f_{\min}$                & $\SI{0.03}{Hz}$ & Lower frequency bound \\
    $f_{\max}$                & $\SI{2.0}{Hz}$  & Upper frequency bound \\
    \bottomrule
  \end{tabular}
\end{table}

% ---------------------------------------------------------------
\section{Comparison with the Aranovskiy Tracker}

Both trackers estimate the dominant frequency sample by sample with $O(1)$
cost, but their internal mechanisms differ fundamentally.

The Aranovskiy tracker identifies $\theta = -\omega^2$
directly via continuous-time gradient descent, then squares to get frequency.
It has no explicit noise model and its convergence rate is amplitude-dependent.

The KalmANF tracker identifies $a = 2\cos\hat\omega_d$ via a Kalman update
on the notch coefficient. The noise model ($q$, $r$) is explicit and
separately tunable from the resonator sharpness ($\rho$). The Kalman structure
gives automatic step-size scheduling, while the Aranovskiy energy-weighting
$\rho_{\rm energy}$ provides a simpler form of the same service.

In practice, KalmANF tends to be more robust when signal amplitude varies
widely, because the Kalman gain already normalises by $s[n-1]^2$ (signal
power at the resonant frequency), providing automatic gain control. The
Aranovskiy tracker's explicit $k$ gain requires re-tuning if the input scale
changes significantly.

% ---------------------------------------------------------------
\section{Numerical Notes}

\subsection{Clamping and Floor Values}

Three hard guards prevent floating-point failures:
\begin{itemize}[leftmargin=1.4em]
  \item $a$ is clamped to $(-1.99999, 1.99999)$ before $\arccos$. This
        prevents domain errors at the cost of a maximum frequency error of
        $\arccos(1.99999/2) - 0 \approx 0.45$~mrad/sample.
  \item $p$ is floored at $10^{-12}$, preventing the gain from producing
        NaN through a $0/0$ division.
  \item The denominator of $K[n]$ has a $10^{-12}$ additive guard for
        the case where $s[n-1] = 0$ and $r/p \approx 0$ simultaneously.
\end{itemize}

\subsection{Precision Requirements}

For $\rho > 0.99$, the resonator poles are close to the unit circle and the
states $s[n]$ can grow large before stabilising. Single-precision
(32-bit) floating point has limited dynamic range in this regime. Double
precision (64-bit) is recommended when $\rho > 0.99$, or when the tracked
frequency is very low relative to $f_s$ (e.g., $\hat{f} < 0.01 f_s$).

% ---------------------------------------------------------------
\section{Summary}

The KalmANF C\texttt{++} implementation is a clean scalar EKF applied to the
adaptive notch coefficient:
\begin{itemize}
  \item one resonator state pair $(s[n-1], s[n-2])$,
  \item one adaptive parameter $a \approx 2\cos\hat\omega_d$,
  \item one Kalman covariance $p$ scheduling the gain automatically,
  \item one frequency output $\hat{f} = \arccos(a/2) / (2\pi\Delta t)$,
  \item a three-component confidence and lock layer for reliability.
\end{itemize}

Convergence to the true frequency is guaranteed under persistent excitation.
The tracker is not inherently DC-robust; a high-pass pre-filter is recommended
for sensor deployments where bias or gravity contamination is expected.

\begin{thebibliography}{1}

\bibitem{AliWaterschoot2023}
R.~Ali and T.~van Waterschoot,
``A frequency tracker based on a {Kalman} filter update of a single parameter
adaptive notch filter,''
in \emph{Proc.\ 26th Int.\ Conf.\ Digital Audio Effects (DAFx23)},
Copenhagen, Denmark, Sep.\ 2023.

\end{thebibliography}

\end{document}
