\documentclass[conference]{IEEEtran}

% ---------------------------------------------------------------
% ENCODING AND FONTS
% ---------------------------------------------------------------
\usepackage{iftex}
\ifPDFTeX
  \usepackage[utf8]{inputenc}
  \usepackage[T1]{fontenc}
\fi

\usepackage{newtxtext,newtxmath}
\usepackage{textcomp}

% ---------------------------------------------------------------
% MATH
% ---------------------------------------------------------------
\usepackage{amsmath,bm}
\usepackage{mathtools}
\usepackage{mathrsfs}
\interdisplaylinepenalty=2500
\allowdisplaybreaks

% ---------------------------------------------------------------
% LAYOUT / TYPOGRAPHY
% ---------------------------------------------------------------
\usepackage{microtype}
\usepackage{balance}
\usepackage{flafter}
\usepackage{placeins}
\usepackage{float}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{etoolbox}

% ---------------------------------------------------------------
% TABLES
% ---------------------------------------------------------------
\usepackage{array,booktabs,makecell,adjustbox}

% ---------------------------------------------------------------
% UNITS
% ---------------------------------------------------------------
\usepackage{siunitx}
\sisetup{detect-weight=true, detect-family=true, per-mode=symbol}

% ---------------------------------------------------------------
% REFERENCES / CROSS-REFERENCES
% ---------------------------------------------------------------
\usepackage{cite}
\usepackage[breaklinks=true,hidelinks]{hyperref}
\usepackage{bookmark}
\usepackage[nameinlink,capitalise,noabbrev]{cleveref}

% ---------------------------------------------------------------
% FLOAT TUNING
% ---------------------------------------------------------------
\setcounter{topnumber}{5}
\setcounter{dbltopnumber}{5}
\renewcommand{\topfraction}{0.95}
\renewcommand{\dbltopfraction}{0.95}
\renewcommand{\textfraction}{0.05}
\renewcommand{\floatpagefraction}{0.9}
\renewcommand{\dblfloatpagefraction}{0.9}

% ---------------------------------------------------------------
% CONVENIENCE MACROS
% ---------------------------------------------------------------
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\fitcol}[1]{\adjustbox{max width=\columnwidth}{\ensuremath{\displaystyle #1}}}

% ---------------------------------------------------------------
% DESCRIPTION LIST STYLE
% ---------------------------------------------------------------
\setlist[description]{font=\footnotesize\bfseries,
                      labelsep=0.6em, leftmargin=1.8em}

% ---------------------------------------------------------------
% EQUATION FONT SIZE
% ---------------------------------------------------------------
\AtBeginEnvironment{equation}{\small}
\AtBeginEnvironment{align}{\small}
\AtBeginEnvironment{alignat}{\small}

% ---------------------------------------------------------------
% PDF BOOKMARK SAFETY
% ---------------------------------------------------------------
\pdfstringdefDisableCommands{%
  \def\SI#1#2{#1 #2}%
  \def\bm#1{#1}%
  \def\sigma{sigma}%
  \def\omega{omega}%
  \def\theta{theta}%
  \def\phi{phi}%
  \def\zeta{zeta}%
  \def\hat#1{#1}%
  \def\cdot{}%
  \def\;{ }%
}

% ---------------------------------------------------------------
% UNICODE PUNCTUATION  (pdfLaTeX only)
% ---------------------------------------------------------------
\ifPDFTeX
  \DeclareUnicodeCharacter{2013}{\textendash}
  \DeclareUnicodeCharacter{2014}{\textemdash}
  \DeclareUnicodeCharacter{2018}{\textquoteleft}
  \DeclareUnicodeCharacter{2019}{\textquoteright}
  \DeclareUnicodeCharacter{201C}{\textquotedblleft}
  \DeclareUnicodeCharacter{201D}{\textquotedblright}
  \DeclareUnicodeCharacter{2212}{\textminus}
\fi

\emergencystretch=2em

% ---------------------------------------------------------------
% TITLE AND AUTHOR
% ---------------------------------------------------------------
\title{Analysis of the Aranovskiy Frequency Estimator:\
Paper-to-Code Correspondence, Theory, and Numerical Considerations}

\author{%
  \IEEEauthorblockN{Mikhail S.\ Grushinskiy}
  \IEEEauthorblockA{Independent Researcher, Fair Lawn, NJ, USA\\
    Bareboat Necessities GitHub Project\\
    \texttt{mgrouch@users.github.com}}%
}

% ---------------------------------------------------------------
\begin{document}
\maketitle

% ---------------------------------------------------------------
\begin{abstract}
This paper documents the Aranovskiy frequency estimator in the original 2013 IFAC paper and in the corresponding C\texttt{++} implementation, making explicit both the paper's convergence argument and the practical code-level modifications. The underlying paper presents a genuine convergence
argument: it starts from a classical adaptive law for the unknown parameter,
proves exponential convergence of a first-order observer state by Lemma~1,
and then reduces the realizable estimator to the classical law up to an
exponentially decaying observer error. However, the shipped implementation is
not identical to the bare paper equations: it adds explicit Euler
integration, internal time scaling, signal-energy normalization, hard
clipping of the adaptation rate, and heuristic confidence and lock outputs.

The main implementation facts are as follows. The observer part is
first-order, not second-order: a single low-pass state is integrated, while a
second dynamic state is the adaptive integrator. The adaptive parameter is
formed algebraically from an integrator state and an instantaneous
cross-term. The code uses a one-step-explicit update ordering in which the
adaptation law uses the stored parameter from the previous sample, while the
reported frequency uses a freshly recomputed algebraic parameter before the
integrator is advanced. The estimator is lightweight and practical for
embedded low-frequency tracking, but sensitivity to DC offsets remains an
important limitation.
\end{abstract}

\begin{IEEEkeywords}
frequency estimation, adaptive observer, Aranovskiy estimator,
nonlinear adaptive control, sinusoidal tracking, embedded signal processing
\end{IEEEkeywords}

% ---------------------------------------------------------------
\section{Introduction and Motivation}\label{sec:intro}

Tracking the frequency of a dominant sinusoid is a standard problem in signal
processing and control. Classical approaches such as phase-locked loops,
zero-crossing detectors, and short-time Fourier methods all trade off
latency, noise sensitivity, or computational cost.

The estimator studied here belongs to the nonlinear adaptive-observer family
and is associated with the work of Bobtsov, Nikolaev, Slita, Borgul, and
Aranovskiy~\cite{BobtsovNikolaevSlitaBorgulAranovskiy2013}. It is attractive
for embedded applications because it operates sample by sample with constant
cost and without FFT buffers.

A recurring source of confusion is that secondary descriptions often mix
three different objects:
\begin{enumerate}[leftmargin=1.5em]
  \item the ideal adaptive law used for the theoretical argument,
  \item the realizable continuous-time estimator written in the paper, and
  \item the practical discrete-time C\texttt{++} implementation.
\end{enumerate}
The purpose of this note is to keep those three layers separate and then map
them carefully onto one another.

% ---------------------------------------------------------------
\section{Signal Model}\label{sec:signal}

The measured scalar signal is modeled as
\begin{equation}\label{eq:signal}
  y(t) = A(t)\,\sin\!\bigl(\omega(t)t + \phi(t)\bigr) + \nu(t),
\end{equation}
where
\begin{description}
  \item[$A(t)$] is the signal amplitude,
  \item[$\omega(t)$] is the unknown angular frequency,
  \item[$\phi(t)$] is the phase,
  \item[$\nu(t)$] is measurement noise or disturbance.
\end{description}

The 2013 paper formulates its main result for the simpler constant-frequency,
constant-amplitude disturbed signal
\begin{equation}\label{eq:paper_signal}
  y(t)=\sigma\sin(\omega t+\varphi)+\delta(t),
\end{equation}
with unknown constants $\sigma$, $\omega$, and $\varphi$, and a bounded smooth
disturbance $\delta(t)$~\cite{BobtsovNikolaevSlitaBorgulAranovskiy2013}.

% ---------------------------------------------------------------
\section{What the Original Paper Actually Proves}\label{sec:paper_theory}

The paper does present a real convergence argument, but it is important to
state precisely what form that argument takes.

\subsection{Ideal Adaptive Law}

For the undisturbed sinusoid, the paper writes the auxiliary model
\begin{align}
  \dot{x}_1 &= x_2, \\
  \dot{x}_2 &= -\omega^2 x_1,
\end{align}
and defines
\begin{equation}
  \theta = -\omega^2.
\end{equation}
It then introduces the ideal adaptive law
\begin{equation}\label{eq:ideal_theta_hat}
  \dot{\hat\theta}(t) = k x_1^2\bigl(\theta-\hat\theta(t)\bigr),
\end{equation}
which the paper explicitly describes as a classical tuning method ensuring
asymptotic convergence of $\hat\theta(t)$ to $\theta$ under the cited
limit-integral nondegeneration conditions from earlier Aranovskiy work
(corresponding to persistent excitation in modern language)
~\cite{BobtsovNikolaevSlitaBorgulAranovskiy2013}.

\subsection{Observer Lemma}

Because $x_1$ is not measured directly, the paper constructs the first-order
observer
\begin{equation}\label{eq:paper_observer}
  \dot{\hat{x}}_1 = -a\hat{x}_1 + by,
\end{equation}
with $a=k_1/k_2$ and $b=k_2^{-1}$, and proves in Lemma~1 that the estimation
error $\tilde{x}_1=x_1-\hat{x}_1$ converges exponentially to zero
~\cite{BobtsovNikolaevSlitaBorgulAranovskiy2013}.

\subsection{Realizable Algorithm in the Paper}

After substituting the observer estimate into the ideal law, eliminating the
unmeasured derivative by means of an auxiliary variable $\zeta$, and
neglecting the exponentially decaying observer error, the paper arrives at
its realizable estimator:
\begin{align}
  \dot{\hat{x}}_1 &= -a\hat{x}_1 + by, \label{eq:paper6}\\
  \dot{\zeta} &= -k\hat{x}_1^2\hat\theta - ka\hat{x}_1\dot{\hat{x}}_1
                 - kb\dot{\hat{x}}_1 y, \label{eq:paper10}\\
  \hat\theta &= \zeta + kb\hat{x}_1 y,
  \qquad
  \hat\omega = \sqrt{|\hat\theta|}. \label{eq:paper11}
\end{align}
The paper then states that \eqref{eq:paper6}, \eqref{eq:paper10}, and
\eqref{eq:paper11} are equivalent to the classical algorithm up to the
exponentially decaying observer error and therefore inherit its stability and
asymptotic convergence properties for the sinusoidal input
~\cite{BobtsovNikolaevSlitaBorgulAranovskiy2013}.

\subsection{What This Does \emph{Not} Mean}

This short conference paper does \emph{not} package the realizable estimator
as a long standalone theorem-proof development for the exact final equations.
Instead, the proof structure is: classical adaptive law $+$ observer lemma
$+$ equivalence argument. That is still a genuine convergence argument, and
it should not be described as ``no proof.''

% ---------------------------------------------------------------
\section{Mapping from Paper Notation to the C\texttt{++} Code}\label{sec:mapping}

The provided implementation follows the same structural estimator as the
paper, with the following correspondence:
\begin{align}
  x_1 &\leftrightarrow \hat{x}_1, \\
  \sigma &\leftrightarrow \zeta, \\
  \theta &\leftrightarrow \hat\theta.
\end{align}
Thus the code keeps one observer state $x_1$, one adaptive integrator state
$\sigma$, and forms the algebraic parameter $\theta$ from
\begin{equation}\label{eq:theta_alg}
  \theta = \sigma + kbx_1y.
\end{equation}
The reported internal angular-frequency magnitude is then
\begin{equation}\label{eq:omega_int}
  \omega_{\mathrm{int}} = \sqrt{\max\!\bigl(|\theta|,10^{-7}\bigr)}.
\end{equation}

Relative to the bare continuous-time paper equations, the code adds:
\begin{itemize}[leftmargin=1.4em]
  \item explicit Euler discretization,
  \item internal time scaling,
  \item signal-energy weighting of the update,
  \item hard clipping of the integrator rate,
  \item heuristic phase, confidence, and lock outputs.
\end{itemize}
Therefore, the literature directly supports the underlying estimator
structure, while the exact source code is a modified discrete-time
engineering realization.

% ---------------------------------------------------------------
\section{Implemented Observer Structure}\label{sec:observer}

\subsection{First-Order Low-Pass State}

A common mistake is to call the implemented observer ``second-order.'' The
observer part is actually \textbf{first-order}:
\begin{equation}\label{eq:x1_ode}
  \dot{x}_1 = -a x_1 + b y.
\end{equation}
There is no independently integrated state $x_2$ in the code.

At the same time, the \emph{overall estimator} has dynamic order two, because
it contains the low-pass state $x_1$ and the adaptive integrator state
$\sigma$. This matches the dynamic-order statement in the paper.

Equation~\eqref{eq:x1_ode} is a first-order low-pass filter with time
constant $1/a$ and DC gain $b/a$. When $a=b$, the DC gain is unity.

\subsection{Filtered-State Interpretation}

For sinusoidal input $y(t)=A\sin(\omega_0 t)$, the steady-state response of
\eqref{eq:x1_ode} is
\begin{equation}\label{eq:x1_sine}
  x_1(t)=\frac{b}{\sqrt{a^2+\omega_0^2}}A
  \sin\!\bigl(\omega_0 t - \arctan(\omega_0/a)\bigr).
\end{equation}
Hence the phase lag is
\begin{equation}
  -\arctan(\omega_0/a).
\end{equation}
At $\omega_0=a$ the lag is exactly $-45^{\circ}$, and it approaches
$-90^{\circ}$ only in the high-frequency limit $\omega_0\gg a$.
Consequently, $x_1$ is not an exact quadrature signal; it is a filtered,
phase-lagged proxy for the input.

% ---------------------------------------------------------------
\section{Internal Time Scaling}\label{sec:timescale}

The implementation introduces a fixed time-scale factor
\begin{equation}
  S = 20.
\end{equation}
If the user-facing parameter is $\omega_{\mathrm{up}}$, the code sets
\begin{align}
  a &= S\,\omega_{\mathrm{up}}, \\
  b &= S\,\omega_{\mathrm{up}}, \\
  k &= S\,k_{\mathrm{user}}.
\end{align}
If the physical sample period is $\Delta t$, the internal Euler step is
\begin{equation}
  \delta t = \frac{\Delta t}{S}.
\end{equation}
Likewise, the user-provided initial values of $\theta$ and $\sigma$ are
stored internally after multiplication by $S^2$.

The physical output frequency in hertz is therefore not
$\omega_{\mathrm{int}}/(2\pi)$, but rather
\begin{equation}\label{eq:f_phys}
  f_{\mathrm{phys}} = \frac{\omega_{\mathrm{int}}}{2\pi S}.
\end{equation}
This matches the source code exactly.

% ---------------------------------------------------------------
\section{Exact Discrete-Time Update Order in the Code}\label{sec:discrete}

Let $y_n$ be the current sample, and let $x_{1,n}$, $\sigma_n$, and
$\theta_n^{\mathrm{stored}}$ denote the values held at the beginning of the
sample. The code performs the following steps.

\subsection{Low-Pass Derivative}

First, it computes
\begin{equation}\label{eq:x1dot_n}
  \dot{x}_{1,n} = -a x_{1,n} + b y_n.
\end{equation}

\subsection{Signal-Energy Weight}

Then it forms
\begin{align}
  E_n &= x_{1,n}^2 + y_n^2 + 10^{-12}, \\
  \rho_n &= \frac{E_n}{E_n + 10^{-6}}.
\end{align}
This factor tends to $1$ for strong signals and toward $0$ when the signal
collapses.

\subsection{Adaptation Error}

The code next computes
\begin{align}
  \mathrm{tmp}_n &= b y_n + a x_{1,n}, \\
  \varphi_n &= \dot{x}_{1,n}\,\mathrm{tmp}_n + x_{1,n}^2\theta_n^{\mathrm{stored}}.
\end{align}
This matches the paper's realizable law structurally, except for the extra
weighting factor $\rho_n$ and later clipping.

\subsection{Integrator Rate}

The adaptive integrator rate is
\begin{equation}\label{eq:sigma_dot_n}
  \dot{\sigma}_n = \mathrm{clip}\!\bigl(-k\varphi_n\rho_n,-10^7,10^7\bigr).
\end{equation}
The clip is a practical numerical safeguard not present in the paper.

\subsection{Fresh Algebraic Parameter and Frequency Output}

Before integrating the dynamic states, the code recomputes
\begin{equation}\label{eq:theta_out}
  \theta_n^{\mathrm{out}} = \sigma_n + k b x_{1,n} y_n.
\end{equation}
The output frequency is then obtained from
\begin{align}
  \omega_{\mathrm{int},n}
  &= \sqrt{\max\!\bigl(|\theta_n^{\mathrm{out}}|,10^{-7}\bigr)}, \\
  f_{\mathrm{phys},n}
  &= \frac{\omega_{\mathrm{int},n}}{2\pi S}.
\end{align}

A crucial detail is that the adaptation error $\varphi_n$ uses the stored
parameter $\theta_n^{\mathrm{stored}}$, while the reported frequency uses the
freshly recomputed algebraic value $\theta_n^{\mathrm{out}}$.

\subsection{Euler Integration}

The dynamic states are advanced by explicit Euler:
\begin{align}
  x_{1,n+1} &= x_{1,n} + \dot{x}_{1,n}\,\delta t, \\
  \sigma_{n+1} &= \sigma_n + \dot{\sigma}_n\,\delta t.
\end{align}
The member variable $\theta$ in the code retains
$\theta_n^{\mathrm{out}}$ for use in the next sample's adaptation law.

\subsection{Phase Heuristic}

Finally, the implementation reports
\begin{equation}\label{eq:phase_est}
  \hat{\phi}_n = \operatorname{atan2}(x_{1,n+1},y_n).
\end{equation}
This is a heuristic phase proxy, not an exact analytic-signal phase.

\begin{table}[t]
  \centering
  \caption{Main variables in the implementation.}
  \label{tab:vars}
  \small
  \begin{tabular}{@{}llp{3.25cm}@{}}
    \toprule
    Symbol & Units & Meaning \\
    \midrule
    $x_1$        & signal     & low-pass state \\
    $\sigma$     & scaled     & adaptive integrator state \\
    $\theta$     & scaled     & algebraic parameter estimate \\
    $\varphi$    & scaled     & adaptation error signal \\
    $\rho$       & $[0,1]$    & signal-energy weight \\
    $\omega_{\mathrm{int}}$ & rad/s scaled & internal angular-frequency magnitude \\
    $f_{\mathrm{phys}}$ & Hz & physical output frequency \\
    $\hat\phi$   & rad        & heuristic phase output \\
    $S$          & ---        & time-scale factor \\
    \bottomrule
  \end{tabular}
\end{table}

% ---------------------------------------------------------------
\section{Connection Between the Paper and the Code}\label{sec:connection}

If one removes the code-specific modifications
\begin{equation}
  \rho_n \to 1,
  \qquad
  \mathrm{clip}(u,-10^7,10^7) \to u,
\end{equation}
and ignores the internal time scaling, then the source code reduces to the
same structural estimator as \eqref{eq:paper6}--\eqref{eq:paper11}. In this
sense the code is not an unrelated algorithm; it is a practical realization
of the same observer-plus-adaptation architecture.

What changes is the scope of the theory one can cite directly. The paper's
convergence argument applies to the continuous-time estimator structure.
The exact C\texttt{++} source code additionally contains discretization,
weighting, clipping, and heuristics. Those changes are reasonable
engineering modifications, but they are not themselves proved in the 2013
paper.

% ---------------------------------------------------------------
\section{Signal Processing Interpretation}\label{sec:dsp}

\subsection{Relation to Quadrature and the Hilbert Viewpoint}

The implementation does not compute the Hilbert transform explicitly. Still,
for a narrowband sinusoid, the filtered state $x_1$ can serve as a useful
lagged companion signal, which is why \eqref{eq:phase_est} can behave like a
rough phase estimate after lock. This interpretation is qualitative rather
than exact.

\subsection{Nyquist Requirement}

As with any sampled-data frequency tracker, the signal of interest must lie
comfortably below the Nyquist frequency
\begin{equation}
  f_N = \frac{1}{2\Delta t}.
\end{equation}
Any significant energy above $f_N$ aliases into the baseband and can drive
the estimator toward the wrong frequency. Internal time scaling does not
remove this requirement.

% ---------------------------------------------------------------
\section{Noise and Bias Sensitivity}\label{sec:noise}

\subsection{Zero-Mean Broadband Noise}

For zero-mean broadband noise, the adaptive law is typically jittered rather
than permanently biased. The low-pass state and the adaptive integrator both
average part of that fluctuation, so the estimator is usually tolerant of
moderate measurement noise.

\subsection{DC Offset and Slow Drift}

The estimator is not inherently DC-robust. If the measurement contains a
constant offset $y_0$, then products such as $x_1 y$ acquire biased terms,
which can shift the parameter estimate and therefore the frequency output.
This is one of the most important practical limitations of the method.

The usual mitigation is to apply mean removal, AC coupling, or a gentle
high-pass filter before frequency estimation.

% ---------------------------------------------------------------
\section{Confidence and Lock Heuristics}\label{sec:lock}

The original paper contains no lock detector. The implementation therefore
adds a synthetic diagnostic layer.

\subsection{Signal RMS}

A one-pole smoother tracks the instantaneous power $x_1^2+y^2$ and produces a
smoothed RMS-like magnitude.

\subsection{Frequency Stability}

A second one-pole smoother tracks the physical frequency estimate. The
instantaneous deviation between the raw and smoothed frequencies is converted
into a stability score.

\subsection{Composite Confidence and Lock}

Amplitude sufficiency and frequency stability are combined into a confidence
score in $[0,1]$, which is then low-pass filtered again. The lock flag is set
only when the frequency lies in a configured range, the signal RMS exceeds a
threshold, and the confidence exceeds its own threshold. These outputs are
helpful engineering indicators, but they are not part of the published
convergence theory.

% ---------------------------------------------------------------
\section{Numerical Implementation Notes}\label{sec:numerical}

\subsection{Initial Conditions}

The paper suggests a practical initialization near
$-\omega_{\mathrm{up}}^2/4$ when an upper bound on frequency is known
~\cite{BobtsovNikolaevSlitaBorgulAranovskiy2013}. That is a sensible starting
point for both $\theta$ and $\sigma$.

However, the exact reason the code always reports a real frequency is not the
sign of the initial condition. It is the output law
\begin{equation}
  \omega_{\mathrm{int}} = \sqrt{\max(|\theta|,10^{-7})},
\end{equation}
which uses an absolute value and a positive floor before the square root.

\subsection{Euler Stability}

For the first-order filter state, explicit Euler is well behaved when the
internal step satisfies the usual small-step condition relative to the pole
$a$. In the intended operating regime of the code, the sample rate is much
higher than the tracked frequency band, so this is normally not restrictive.

\subsection{Overflow Guards}

Two important safeguards are built in:
\begin{itemize}[leftmargin=1.4em]
  \item clipping of $\dot\sigma$ to $[-10^7,10^7]$,
  \item flooring of $|\theta|$ before the square root.
\end{itemize}
These are practical protections for finite-precision embedded arithmetic.

\begin{table}[t]
  \centering
  \caption{Main tuning parameters and their roles.}
  \label{tab:params}
  \small
  \begin{tabular}{@{}lp{2.0cm}p{2.7cm}@{}}
    \toprule
    Parameter & Role & Practical effect \\
    \midrule
    $a$ & low-pass pole & larger $a$ gives faster state response and less lag \\
    $b$ & input gain & sets the observer gain together with $a$ \\
    $k$ & adaptation gain & larger $k$ increases pull of the parameter update \\
    $S$ & time scaling & improves low-frequency numerical conditioning \\
    $\rho$ logic & weak-signal protection & suppresses adaptation when energy is tiny \\
    clipping & safety guard & prevents integrator runaway on transients \\
    \bottomrule
  \end{tabular}
\end{table}

% ---------------------------------------------------------------
\section{Summary}\label{sec:summary}

The correct high-level picture is as follows.

First, the 2013 Aranovskiy paper does provide a convergence argument: it uses
an ideal classical adaptive law, proves exponential convergence of the
observer state by Lemma~1, and then transfers that result to the realizable
estimator through an equivalence argument.

Second, the C\texttt{++} implementation is structurally faithful to that
realizable estimator, with the paper variables mapping naturally to the code
states and algebraic quantities.

Third, the exact source code is still a modified engineering realization, not
a verbatim copy of the continuous-time equations. Its most important added
features are internal time scaling, signal-energy weighting, clipped
adaptation, and heuristic confidence/lock outputs.

Finally, the observer portion of the implementation is first-order, while the
overall estimator remains second-order because it consists of one low-pass
state and one adaptive integrator. That distinction resolves the main
paper-to-code confusion.

% ---------------------------------------------------------------
\section*{Acknowledgments}
This document was prepared by M.~S.~Grushinskiy as technical documentation
for the Bareboat Necessities open-source project.

% ---------------------------------------------------------------
\begin{thebibliography}{1}
\bibitem{BobtsovNikolaevSlitaBorgulAranovskiy2013}
A.\,A.~Bobtsov, N.\,A.~Nikolaev, O.\,V.~Slita, A.\,S.~Borgul, and
S.\,V.~Aranovskiy,
``The new algorithm of sinusoidal signal frequency estimation,''
in \emph{Proc. 11th IFAC Int. Workshop Adaptation and Learning in Control
and Signal Processing}, Caen, France, Jul. 2013, pp. 182--185.
\end{thebibliography}

\balance
\end{document}
