\documentclass[conference]{IEEEtran}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{newtxtext,newtxmath}
\usepackage{amsmath,bm}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{enumitem}
\usepackage[breaklinks=true,hidelinks]{hyperref}

\title{A Practical Description of the Adaptive Vertical PII Observer,\\
Its Mahony Wrapper, and Runtime Adaptation}

\author{%
  \IEEEauthorblockN{Mikhail S.\ Grushinskiy}
  \IEEEauthorblockA{Independent Researcher\\
  Bareboat Necessities Project\\
  \texttt{mgrouch@users.github.com}}
}

\begin{document}
\maketitle

\begin{abstract}
This note gives a practical description of a lightweight vertical-motion
observer stack for embedded marine use. The stack has three layers.

First, a vertical PII observer estimates displacement and velocity from
world-frame vertical inertial acceleration using a repeated-real-pole core
with an optional very-slow bias channel. Second, a Mahony-based wrapper
estimates tilt, rotates body-frame specific force into a world \(Z\)-up frame,
and extracts vertical inertial acceleration. Third, an adaptive scheduling
layer uses an external or internal frequency proxy together with an
acceleration-level estimate to tune the observer conservatively at runtime.

The document follows the code structure closely and focuses on what the
implementation actually computes, how the layers connect, and how the
simulation-backed tuning values are intended to be interpreted.
\end{abstract}

\begin{IEEEkeywords}
heave estimation, vertical observer, PII observer, Mahony AHRS, adaptive
scheduling, marine inertial sensing
\end{IEEEkeywords}

% ---------------------------------------------------------------
\section{Introduction}

The implementation is organized as a small stack.

\begin{enumerate}[leftmargin=1.5em]
  \item \textbf{VerticalPIIObserver}: the core vertical-motion observer.
  \item \textbf{AdaptiveVerticalPII}: a wrapper that adds runtime scheduling
        and a policy-selected frequency tracker.
  \item \textbf{AdaptiveVerticalPIIMahony}: a wrapper that uses Mahony AHRS
        to obtain the vertical inertial acceleration fed into the adaptive
        vertical observer.
\end{enumerate}

The intended signal path is:
\begin{equation}
\text{body IMU} \rightarrow \text{Mahony tilt} \rightarrow
a_{z,\mathrm{world,up}} \rightarrow \text{adaptive vertical observer}
\rightarrow (p,v).
\end{equation}

The code is deliberately simple and embedded-friendly:
no dynamic allocation, no matrix library, and only scalar update equations.

% ---------------------------------------------------------------
\section{Vertical PII Observer}\label{sec:pii}

\subsection{Purpose and States}

The core observer estimates vertical displacement and velocity from an input
that is already the gravity-compensated world-frame vertical inertial
acceleration.

Its main states are:
\begin{align}
a_f &: \text{filtered acceleration},\\
v   &: \text{vertical velocity},\\
p   &: \text{vertical displacement},\\
S   &: \text{integral of displacement}.
\end{align}

When the optional bias channel is enabled, it also keeps
\begin{align}
d &: \text{very-slow displacement trend},\\
\hat b &: \text{estimated acceleration bias}.
\end{align}

\subsection{Core Equations}

The non-oscillatory observer core is
\begin{align}
\dot a_f &= \frac{a_{\mathrm{meas}} - a_f}{\tau_a}, \\
\dot S   &= p, \\
\dot p   &= v, \\
\dot v   &= a_f - \hat b - k_v v - k_p p - k_s S.
\end{align}

So the measured acceleration first passes through a low-pass stage, and then
drives a third-order restoring observer acting on \(v\), \(p\), and the
integral state \(S\).

\subsection{Repeated-Real-Pole PII Parameterization}

The base PII core is parameterized by a repeated real pole \(r>0\):
\begin{equation}
(s+r)^3 = s^3 + 3rs^2 + 3r^2 s + r^3.
\end{equation}

The implementation therefore sets
\begin{equation}
k_v = 3r, \qquad
k_p = 3r^2, \qquad
k_s = r^3.
\end{equation}

This is an important design choice. Instead of tuning three restoring gains
independently, the code uses one motion-pole-rate parameter \(r\), then
derives the PII gains from the repeated-pole relation.

\subsection{Optional Very-Slow Bias Channel}

When \texttt{WithBias=true}, the observer adds
\begin{align}
\dot d     &= \frac{p-d}{\tau_d},\\
\dot{\hat b} &= k_b d - \lambda_b \hat b.
\end{align}

Here \(d\) is a very slow trend extracted from displacement, and \(\hat b\) is
a leaky estimate of a residual acceleration bias. The bias estimate is then
subtracted inside \(\dot v\).

This channel is conservative by design: it is intended to handle very slow
bias drift, not to follow fast wave dynamics.

\subsection{Per-Sample Update}

The implementation uses a one-pole coefficient
\begin{equation}
\alpha(dt,\tau) = \frac{dt}{\tau + dt},
\end{equation}
not an exponential form.

At each sample:
\begin{enumerate}[leftmargin=1.5em]
  \item the filtered acceleration is updated,
        \[
        a_f \leftarrow a_f + \alpha(dt,\tau_a)(a_{\mathrm{meas}}-a_f),
        \]
  \item if the bias channel is enabled, \(d\) and \(\hat b\) are updated,
  \item the vertical acceleration model
        \[
        \dot v = a_f - \hat b - k_v v - k_p p - k_s S
        \]
        is formed,
  \item a semi-implicit state update is applied:
        \begin{align}
        v &\leftarrow v + dt\,\dot v,\\
        p &\leftarrow p + dt\,v,\\
        S &\leftarrow S + dt\,p.
        \end{align}
\end{enumerate}

The use of the updated \(v\) inside the position update, and the updated \(p\)
inside the integral update, matches the code exactly.

\subsection{State and Safety Clamps}

The implementation supports optional symmetric clamps on:
\begin{equation}
a_f,\quad v,\quad p,\quad S,\quad d,\quad \hat b.
\end{equation}

A non-positive limit disables the corresponding clamp. These are simply
practical safety bounds for embedded use and pathological transients.

\begin{table}[t]
  \centering
  \caption{Main states of the vertical PII observer}
  \label{tab:pii_states}
  \small
  \begin{tabular}{@{}lll@{}}
    \toprule
    Quantity & Code field & Meaning \\
    \midrule
    \(a_f\) & \texttt{a\_f\_} & filtered vertical acceleration \\
    \(v\)   & \texttt{v\_}    & vertical velocity \\
    \(p\)   & \texttt{p\_}    & vertical displacement \\
    \(S\)   & \texttt{S\_}    & integral of displacement \\
    \(d\)   & \texttt{bias\_.d} & slow displacement trend \\
    \(\hat b\) & \texttt{bias\_.b} & estimated accel bias \\
    \(r\)   & \texttt{active\_r\_} & active repeated pole rate \\
    \bottomrule
  \end{tabular}
\end{table}

% ---------------------------------------------------------------
\section{Mahony Wrapper}\label{sec:mahony}

\subsection{Role of the Wrapper}

The Mahony wrapper is responsible for converting raw body-frame IMU data into
the quantity expected by the vertical observer:
\begin{equation}
a_{z,\mathrm{world,up}} = \text{vertical inertial acceleration, world \(Z\)-up}.
\end{equation}

So the PII observer itself does not estimate attitude. It assumes the input is
already vertical inertial acceleration in a consistent world-frame sign
convention.

\subsection{World-Frame Convention}

The supplied Mahony implementation is effectively a \(Z\)-up world frame.
At rest and identity quaternion, the expected accelerometer reading is
approximately \((0,0,+g)\), so the world ``up'' direction is \(+Z\).

The wrapper therefore computes:
\begin{equation}
a_{\mathrm{world,up}} =
\bigl(R_{bw}\,a_{\mathrm{body}}\bigr)_z - g,
\end{equation}
where:
\begin{itemize}[leftmargin=1.4em]
  \item \(a_{\mathrm{body}}\) is body-frame specific force,
  \item \(R_{bw}\) is the body-to-world rotation,
  \item the subtracted \(g\) is the configured gravity magnitude.
\end{itemize}

This is exactly the vertical signal fed into the adaptive vertical observer.

\subsection{Quaternion Use}

The internal Mahony quaternion is stored as world-to-body. The wrapper uses
its conjugate as body-to-world for vector rotation.

The code performs the vector rotation with the efficient quaternion-vector
formula rather than by constructing a dense rotation matrix.

\subsection{IMU Update Path}

For each IMU sample:
\begin{enumerate}[leftmargin=1.5em]
  \item Mahony gains may be adapted from accelerometer trust,
  \item Mahony is updated either with IMU-only data or with IMU+magnetometer,
  \item the body-frame accelerometer is rotated into the Mahony world frame,
  \item vertical inertial acceleration is extracted as \(a_{z,\mathrm{world,up}}\),
  \item that signal is passed into the adaptive vertical observer.
\end{enumerate}

The wrapper exposes roll, pitch, and yaw in degrees as reported by the Mahony
update functions, together with the latest rotated world-frame acceleration.

% ---------------------------------------------------------------
\section{Adaptive Scheduling}\label{sec:adaptive}

\subsection{Observer-Level Scheduling Variables}

The core observer supports conservative runtime scheduling based on:
\begin{itemize}[leftmargin=1.4em]
  \item a representative displacement frequency \(f_{\mathrm{disp}}\) in Hz,
  \item an acceleration scale \(\sigma_a\),
  \item a confidence value in \([0,1]\).
\end{itemize}

The intended interpretation is:
\begin{itemize}[leftmargin=1.4em]
  \item \(f_{\mathrm{disp}}\) is a scheduling hint, not ground truth,
  \item larger \(\sigma_a\) means rougher motion and less aggressive restoring,
  \item low confidence means ``hold the current parameters.''
\end{itemize}

\subsection{Input Smoothing}

When \texttt{update\_adaptation()} is called with valid values and sufficient
confidence, the observer first smooths the scheduling signals:
\begin{align}
f_{\mathrm{filt}} &\leftarrow
f_{\mathrm{filt}} + \alpha(dt,\tau_{\mathrm{in}})
(f_{\mathrm{disp}} - f_{\mathrm{filt}}),\\
\sigma_{\mathrm{filt}} &\leftarrow
\sigma_{\mathrm{filt}} + \alpha(dt,\tau_{\mathrm{in}})
(\sigma_a - \sigma_{\mathrm{filt}}).
\end{align}

If confidence is below the configured threshold, or if the inputs are invalid,
the active parameters are simply held.

\subsection{Normalized Ratios}

The scheduling logic forms
\begin{equation}
r_f = \frac{f_{\mathrm{filt}}}{f_{\mathrm{ref}}},
\qquad
r_\sigma = \frac{\sigma_{\mathrm{ref}}}{\sigma_{\mathrm{filt}}},
\end{equation}
with mild clamping before exponentiation.

The main scheduled quantities are then built from conservative power laws.

\subsection{Scheduled Motion Pole and Acceleration Filter}

The active repeated pole rate is scheduled as
\begin{equation}
r_{\mathrm{cmd}}
=
r_{\mathrm{base}}\,
r_f^{\,e_{r,f}}\,
r_\sigma^{\,e_{r,\sigma}},
\end{equation}
then clamped to \([r_{\min}, r_{\max}]\).

Similarly, the acceleration filter time constant is scheduled as
\begin{equation}
\tau_{a,\mathrm{cmd}}
=
\tau_{a,\mathrm{base}}\,
r_f^{\,e_{\tau_a,f}}\,
r_\sigma^{\,e_{\tau_a,\sigma}},
\end{equation}
then clamped to \([\tau_{a,\min}, \tau_{a,\max}]\).

With the default exponent signs used in the code:
\begin{itemize}[leftmargin=1.4em]
  \item faster displacement waves allow somewhat faster observer dynamics,
  \item larger acceleration sigma slows the observer slightly and increases
        acceleration smoothing.
\end{itemize}

\subsection{Scheduled Bias Channel}

When the bias channel is enabled, the same logic also schedules
\(\tau_d\) and \(k_b\):
\begin{align}
\tau_{d,\mathrm{cmd}}
&=
\tau_{d,\mathrm{base}}\,
r_f^{\,e_{\tau_d,f}}\,
r_\sigma^{\,e_{\tau_d,\sigma}},\\
k_{b,\mathrm{cmd}}
&=
k_{b,\mathrm{base}}\,
r_f^{\,e_{k_b,f}}\,
r_\sigma^{\,e_{k_b,\sigma}}.
\end{align}

These are then clamped to configured bounds and smoothed before becoming the
active parameters.

\subsection{Active-Parameter Smoothing}

The active parameters do not jump immediately to their scheduled targets.
Instead, the implementation applies another one-pole smoothing stage:
\begin{equation}
\theta_{\mathrm{act}}
\leftarrow
\theta_{\mathrm{act}} + \alpha(dt,\tau_{\mathrm{param}})
(\theta_{\mathrm{cmd}} - \theta_{\mathrm{act}}),
\end{equation}
for each scheduled parameter.

So the scheduling layer is deliberately slow and conservative.

% ---------------------------------------------------------------
\section{AdaptiveVerticalPII Wrapper}\label{sec:adaptive_wrapper}

\subsection{Purpose}

The class \texttt{AdaptiveVerticalPII} combines:
\begin{itemize}[leftmargin=1.4em]
  \item a \texttt{VerticalPIIObserver},
  \item a policy-selected acceleration-frequency tracker,
  \item a running estimate of raw-input acceleration sigma,
  \item logic for automatic or external scheduling updates.
\end{itemize}

The frequency-tracker backend is selected by the template parameter
\texttt{TrackerType}. The code supports multiple tracker families behind a
common policy interface.

\subsection{Acceleration Sigma Estimation}

On every IMU-rate update, the wrapper computes a slow mean and a shorter-time
variance estimate from the raw acceleration input:
\begin{align}
\mu_a &\leftarrow \mu_a + \alpha(dt,\tau_\mu)(a - \mu_a),\\
e_a &= a - \mu_a,\\
\sigma_a^2 &\leftarrow \sigma_a^2 + \alpha(dt,\tau_\sigma)(e_a^2 - \sigma_a^2).
\end{align}

The reported sigma is
\begin{equation}
\sigma_a = \sqrt{\sigma_a^2},
\end{equation}
with a configured floor for robustness.

This sigma estimate is used as the roughness input to the observer scheduler.

\subsection{Acceleration-Frequency Tracker Backend}

The wrapper also steps its selected acceleration-frequency tracker on every
sample. The tracker may provide:
\begin{itemize}[leftmargin=1.4em]
  \item a main frequency estimate,
  \item a raw frequency estimate,
  \item a confidence score,
  \item a lock flag,
  \item optionally a coarse estimate.
\end{itemize}

The wrapper can then use that tracker output as an automatic fallback
scheduling signal when no external displacement-frequency estimate is
available.

\subsection{Preferred and Fallback Scheduling Paths}

The preferred adaptation source is an externally estimated displacement
frequency:
\begin{equation}
\texttt{updateAdaptationFromDisplacementFrequency}(f_{\mathrm{disp}},dt,c).
\end{equation}

There is also a full external path where both frequency and sigma are supplied
explicitly.

If automatic scheduling from the acceleration-frequency proxy is enabled, the
wrapper accumulates time and periodically updates the observer scheduler using:
\begin{itemize}[leftmargin=1.4em]
  \item a scheduled frequency built from the internal tracker,
  \item the internally estimated \(\sigma_a\),
  \item a confidence value derived from tracker confidence, lock, and coarse
        estimate availability.
\end{itemize}

\subsection{Scheduled Frequency from the Internal Tracker}

The internal scheduling frequency is built as follows:
\begin{enumerate}[leftmargin=1.5em]
  \item prefer the tracker’s main frequency estimate,
  \item if invalid, fall back to the tracker’s raw frequency estimate,
  \item if a coarse estimate exists, blend toward it,
  \item blend more strongly toward the coarse estimate when the tracker is not
        locked.
\end{enumerate}

So the auto-scheduled frequency is intentionally more conservative than just
using the raw tracker output directly.

\subsection{Fallback Confidence Used for Scheduling}

The auto-scheduling confidence starts from the tracker’s own confidence.
It is then floored upward when:
\begin{itemize}[leftmargin=1.4em]
  \item a coarse estimate exists,
  \item the tracker is locked.
\end{itemize}

This means the vertical observer can continue adapting moderately even when the
acceleration-frequency proxy is imperfect, as long as it is sufficiently
coarse-valid or locked.

\begin{table}[t]
  \centering
  \caption{Main adaptation quantities}
  \label{tab:adapt}
  \small
  \begin{tabular}{@{}lll@{}}
    \toprule
    Quantity & Code field & Role \\
    \midrule
    \(f_{\mathrm{filt}}\) & \texttt{f\_disp\_filt\_hz\_} & filtered scheduling frequency \\
    \(\sigma_{\mathrm{filt}}\) & \texttt{sigma\_a\_filt\_} & filtered scheduling sigma \\
    \(r_{\mathrm{act}}\) & \texttt{active\_r\_} & active repeated pole rate \\
    \(\tau_{a,\mathrm{act}}\) & \texttt{active\_tau\_a\_} & active accel LPF tau \\
    \(\tau_{d,\mathrm{act}}\) & \texttt{active\_tau\_d\_} & active trend tau \\
    \(k_{b,\mathrm{act}}\) & \texttt{active\_kb\_} & active bias gain \\
    \bottomrule
  \end{tabular}
\end{table}

% ---------------------------------------------------------------
\section{Mahony Gain Adaptation}\label{sec:mahony_adapt}

\subsection{Trust Metrics}

The Mahony wrapper can also adapt Mahony’s gains from accelerometer trust.
It computes three ingredients.

First, a norm-consistency error:
\begin{equation}
e_n = \frac{\bigl|\|a\| - g\bigr|}{g}.
\end{equation}

Second, an innovation magnitude comparing the measured normalized
accelerometer direction with the gravity direction predicted by the current
Mahony quaternion:
\begin{equation}
e_i = \left\|\hat a \times \hat g_{\mathrm{pred}}\right\|.
\end{equation}

Third, it reuses the vertical observer’s current acceleration sigma estimate
\(\sigma_a\).

\subsection{Trust Score}

The code normalizes those quantities against configured reference scales and
forms
\begin{equation}
\rho_n = \frac{1}{1+n^2},\qquad
\rho_i = \frac{1}{1+u^2},\qquad
\rho_\sigma = \frac{1}{1+s^2},
\end{equation}
where \(n\), \(u\), and \(s\) are the normalized norm error, innovation, and
sigma.

The combined trust score is
\begin{equation}
\tau_{\mathrm{acc}} = \rho_n\,\rho_i\,\rho_\sigma,
\end{equation}
then clamped to
\begin{equation}
\tau_{\mathrm{acc}} \in [\tau_{\min},1].
\end{equation}

So accelerometer trust is highest when:
\begin{itemize}[leftmargin=1.4em]
  \item \(\|a\|\) is close to \(g\),
  \item measured and predicted gravity directions agree,
  \item the vertical observer reports calm motion.
\end{itemize}

\subsection{Gain Mapping}

The proportional Mahony gain is scheduled linearly:
\begin{equation}
2K_p =
(2K_p)_{\mathrm{rough}}
+
\tau_{\mathrm{acc}}
\Bigl((2K_p)_{\mathrm{calm}} - (2K_p)_{\mathrm{rough}}\Bigr).
\end{equation}

The integral gain is scheduled more cautiously:
\begin{equation}
2K_i =
(2K_i)_{\mathrm{rough}}
+
\tau_{\mathrm{acc}}^2
\Bigl((2K_i)_{\mathrm{calm}} - (2K_i)_{\mathrm{rough}}\Bigr).
\end{equation}

So the integral term collapses faster than the proportional term as trust
drops.

\subsection{Gain Smoothing}

The commanded Mahony gains are smoothed with the same embedded-friendly
one-pole form:
\begin{equation}
g_{\mathrm{act}}
\leftarrow
g_{\mathrm{act}} + \alpha(dt,\tau_g)(g_{\mathrm{cmd}} - g_{\mathrm{act}}).
\end{equation}

This avoids abrupt gain jumps during rough motion.

% ---------------------------------------------------------------
\section{Simulation-Backed Tuning Example}\label{sec:tuning}

The supplied simulation adapter instantiates
\begin{equation}
\texttt{AdaptiveVerticalPIIMahony<float,true,TrackerType::ARANOVSKIY>}.
\end{equation}

So the adaptive vertical core uses the Mahony wrapper, enables the bias
channel, and uses the Aranovskiy tracker as the internal acceleration-frequency
fallback.

\subsection{Core Vertical Observer Tuning}

The example tuning sets:
\begin{align}
r &= 0.150,\\
\tau_a &= 0.68~\si{s},\\
\tau_d &= 49.0~\si{s},\\
k_b &= 2.5\times 10^{-5},\\
\lambda_b &= 3.0\times 10^{-3}.
\end{align}

Safety limits are
\begin{equation}
|a_f|\le 50,\quad |v|\le 50,\quad |p|\le 20,\quad |S|\le 200,\quad |d|\le 20,
\end{equation}
with
\begin{equation}
|\hat b| \le 0.12.
\end{equation}

The adaptation layer uses:
\begin{align}
f_{\mathrm{ref}} &= 0.12~\si{Hz},\\
\sigma_{\mathrm{ref}} &= 0.95,\\
\tau_{\mathrm{in}} &= 4.5~\si{s},\\
\tau_{\mathrm{param}} &= 7.5~\si{s}.
\end{align}

The example also enables automatic scheduling from the internal
acceleration-frequency proxy every
\begin{equation}
0.50~\si{s}.
\end{equation}

\subsection{Mahony Tuning}

The simulation-backed Mahony tuning uses:
\begin{align}
(2K_p)_{\mathrm{base}} &= 0.45,\\
(2K_i)_{\mathrm{base}} &= 0.015,\\
(2K_p)_{\mathrm{calm}} &= 0.90,\\
(2K_p)_{\mathrm{rough}} &= 0.35,\\
(2K_i)_{\mathrm{calm}} &= 0.025,\\
(2K_i)_{\mathrm{rough}} &= 0.010.
\end{align}

The trust references are
\begin{align}
\sigma_{\mathrm{ref}}^{\mathrm{Mahony}} &= 0.18~\si{m/s^2},\\
e_n^{\mathrm{ref}} &= 0.08,\\
e_i^{\mathrm{ref}} &= 0.12,
\end{align}
with gain-schedule smoothing time constant
\begin{equation}
\tau_g = 2.0~\si{s},
\end{equation}
and minimum trust
\begin{equation}
\tau_{\min} = 0.05.
\end{equation}

\subsection{Frame Mapping Used in the Simulation}

The simulation runner works in body NED. The adapter converts that to the
Mahony body convention by
\begin{equation}
(x,y,z)_{\mathrm{NED}}
\mapsto
(x,-y,-z)_{\mathrm{Mahony}},
\end{equation}
so that gravity is \(+Z\) in Mahony’s frame.

For RMS reporting back in the simulation convention, the example adapter also
maps Mahony Euler outputs back to the nautical convention before comparing
against reference attitudes.

\subsection{Quality Gates Used in the Example}

The simulation code checks the last \SI{60}{s} of each run and applies quality
gates:
\begin{itemize}[leftmargin=1.4em]
  \item Z RMS \(< 17.9\%\) of \(H_s\) for JONSWAP cases,
  \item Z RMS \(< 16.0\%\) of \(H_s\) for PM-Stokes cases,
  \item yaw RMS \(< 5.06^\circ\) when magnetometer is used.
\end{itemize}

These gates are not part of the observer itself, but they are useful context
for the intended tuning regime.

\begin{table}[t]
  \centering
  \caption{Simulation-backed example tuning}
  \label{tab:tuning}
  \small
  \begin{tabular}{@{}lll@{}}
    \toprule
    Parameter & Value & Role \\
    \midrule
    \(r\) & 0.150 & base repeated pole rate \\
    \(\tau_a\) & \SI{0.68}{s} & accel LPF time constant \\
    \(\tau_d\) & \SI{49}{s} & slow trend time constant \\
    \(k_b\) & \(2.5\times10^{-5}\) & bias learning gain \\
    \(\lambda_b\) & \(3.0\times10^{-3}\) & bias leak \\
    \(f_{\mathrm{ref}}\) & \SI{0.12}{Hz} & displacement-frequency reference \\
    \(\sigma_{\mathrm{ref}}\) & 0.95 & observer sigma reference \\
    \((2K_p)_{\mathrm{base}}\) & 0.45 & Mahony base proportional gain \\
    \((2K_i)_{\mathrm{base}}\) & 0.015 & Mahony base integral gain \\
    \(g\) & \SI{9.80665}{m/s^2} & gravity magnitude \\
    \bottomrule
  \end{tabular}
\end{table}

% ---------------------------------------------------------------
\section{Summary}

The code implements a layered vertical-motion estimation stack.

First, the \texttt{VerticalPIIObserver} is a repeated-real-pole PII observer
with optional very-slow bias estimation. It is intentionally conservative,
non-oscillatory, and cheap to run.

Second, the \texttt{AdaptiveVerticalPII} wrapper adds acceleration sigma
estimation, a policy-selected frequency tracker, and runtime scheduling of the
observer parameters. External displacement-frequency scheduling is preferred,
while acceleration-frequency scheduling serves as a fallback.

Third, the \texttt{AdaptiveVerticalPIIMahony} wrapper uses Mahony AHRS to
supply the required vertical inertial acceleration and can also adapt Mahony’s
gains from accelerometer trust and sea-state roughness indicators.

Together, these layers provide a compact embedded heave-estimation pipeline
that is practical to tune, robust to moderate motion variability, and
compatible with several lightweight frequency-tracker backends.

\end{document}
